---
title: "Cellular response to Gemcitabine and other drugs"
author: "Diego Montero Solano, Sophie Zhu, Tabita Häßner, Shivani Nundoo"
date: "6/5/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading the data

```{r}
#Set wd to the working directory where this Rscript is located
wd = getwd()

#Load data
basalexpression <- readRDS(paste0(wd,"/CCLE_basalexpression.rds"))
copynumber <- readRDS(paste0(wd,"/CCLE_copynumber.rds"))
mutations <- readRDS(paste0(wd,"/CCLE_mutations.rds"))
treated <- readRDS(paste0(wd,"/NCI_TPW_gep_treated.rds"))
untreated <- readRDS(paste0(wd,"/NCI_TPW_gep_untreated.rds"))
drugsensitivity <- readRDS(paste0(wd,"/NegLogGI50.rds"))

cannotation <- read.table(paste0(wd,"/cellline_annotation.tsv"), header = TRUE, sep ="\t", stringsAsFactors = TRUE)
dannotation <- read.table(paste0(wd,"/drug_annotation.tsv"), header = TRUE, sep ="\t", stringsAsFactors = TRUE)
metadata <- read.table(paste0(wd,"/NCI_TPW_metadata.tsv"), header = TRUE, sep ="\t", stringsAsFactors = TRUE)
```

## Normalization of data

```{r}
#loading library limma needed for normalization
library(limma)

#normalize treated & untreated
untreated_normalized <- normalizeBetweenArrays(untreated)
treated_normalized <- normalizeBetweenArrays(treated)
```

## Comparing boxplots

```{r}
boxplot(treated, main = "geneexpression after treatment", ylab = "geneexpression", xlab = "samples", names = FALSE)
boxplot(treated_normalized, main = "normalized geneexpression after treatment", ylab = "geneexpression", xlab = "samples", names = FALSE) 
```


## PCA

```{r}
#PCA of FC 
fcgeneexpression <- treated_normalized - untreated_normalized
pca_FC = prcomp(fcgeneexpression, center = F, scale. = F)

#plot PC 1-4 of FC

drug_as_factor <- as.factor(metadata[1:ncol(treated),"drug"])
levels <- as.factor(levels(drug_as_factor))

par(xpd=T, mar=par()$mar+c(0,0,0,6))
plot(pca_FC$rotation[, 1], pca_FC$rotation[, 2], pch = 20, xlab = "PC 1",ylab = "PC 2", col=drug_as_factor, main = "PC 1&2 of the fold change ")
legend("right",inset = c(-0.3,0), levels(drug_as_factor), xpd = TRUE, pch=20, col = levels)

par(xpd=T, mar=par()$mar+c(0,0,0,5))
plot(pca_FC$rotation[, 3], pca_FC$rotation[, 4], pch = 20, xlab = "PC 3",ylab = "PC 4", col=drug_as_factor, main = "PC 3&4 of the fold change ")
legend("right", inset=c(-0.5,0),levels(drug_as_factor), pch=20, col = levels, xpd =TRUE)
```

FC is obviously drug associated

## T-test

Check if normalized treated data is normally distributed using QQ-plot
If normally distributed quantile dots should lie on a straight line 

```{r}
#check if normalized treated data is normally distributed using QQ-plot
qqnorm(treated_normalized, main = "QQ Plot of geneexpression after treatment")
qqline(treated_normalized)
```

T-test with normalized data? Significant difference between treated and untreated?

```{r}
t.test(treated, untreated, paired = TRUE)
t.test(treated_normalized, untreated_normalized, paired = TRUE)
```

The H0 Hypothesis is that there is no change in the mean of the treated and untreated data.

The 95 percent confidence interval is between -0.0015656142 and -0.0009451127 with a p-value of 2.182e-15.
The p-value is not within the 95 percent confidence interval, thus the null hypothesis is rejected for the alternative hypothesis.

Alternative Hypothesis: there is a difference in mean between the treated and untreated data. ("Tentatively": does this prove that the change is provoked by the drug? or is it only a batch effect? Check the boxplots for batch effect identification.

## Biomarkers

Fold change to find the difference in gene expression between treated and untreated celllines 

```{r}
fcgeneexpression <- treated_normalized - untreated_normalized

#Absolute value of the data in fcgeneexpression, to find the biggest changes, regardless of if they are up- or down-regulated
absvalfcgeneexpression <- abs(fcgeneexpression)
```

New matrices with only the celllines treated wiht gemcitabine

```{r}
library("dplyr")

treated_gemcitabine <- select(as_tibble(treated_normalized), contains("gemcitibine"))
treated_gemcitabine <- as.matrix(treated_gemcitabine)
rownames(treated_gemcitabine) <- rownames(treated_normalized)

untreated_gemcitabine <- select(as_tibble(untreated_normalized), contains("gemcitibine"))
untreated_gemcitabine <- as.matrix(untreated_gemcitabine)
rownames(untreated_gemcitabine) <- rownames(untreated_normalized)

absvalfcge_gemcitabine <- select(as_tibble(absvalfcgeneexpression), contains("gemcitibine"))
absvalfcge_gemcitabine <- as.matrix(absvalfcge_gemcitabine)
rownames(absvalfcge_gemcitabine) <- rownames(absvalfcgeneexpression)
```


Create a function to order the components of a vector (descending) and return the names in that order

```{r}
orderforvectors <- function(x) {
  y <- names(x[order(-x)])
  return(y)
}
```

Apply the new funtion *orderforvectors* to all the columns and create a new matrix with the gene names

```{r}
matrixofbiomarkers <- apply(absvalfcge_gemcitabine, 2, orderforvectors) 
#the 2 means it is applied to columns
```

Check for most common genes in the first 15 rows of the new matrix

```{r}
topgenes <- sort(table(as.vector(matrixofbiomarkers[1:15,])))
```

Convert *topgenes* into a matrix in descending order of counts

```{r}
matrixtopgenes <- as.matrix(topgenes[order(-topgenes)])
```

Bar plot of *matrixtopgenes* to find the biomarkers

```{r}
barplot(matrixtopgenes[1:20,], xlab = "", ylab = "Counts", main = "Top differentially expressed genes", cex.names = 0.7, las = 2)
mtext("Genes", side = 1, line = 4)
```
We decided to take the top 15 genes as our biomarkers

List of our biomarkers:

```{r}
biomarkers <- names(matrixtopgenes[1:15,1])
biomarkers
```
## Microsatellite instability

```{r}
# Plotting the GI50 values relating to Gemcitabine treatment .

plot(drugsensitivity["gemcitibine",])

#creating a new dataframe with the Celllines and whether the Microsatellite Instablility is high or not. TRUE:High  False:Stable/Low

L=cannotation$Microsatellite_instability_status=="MSI-H"
Cell_Line=data.frame("Cell Line name"= cannotation[1:61,1],"Microsatellite"= L)
Cell_Line <- Cell_Line[order(Cell_Line$Cell.Line.name),]
drugsensitivity=t(drugsensitivity)

# It is observed that the cell lines in the drugsensitivity matrix is already arranged alphabetically.
# Thus the Cell_Line dataframe is directly merged with the Gemcitabine Drug Sensitivity Values.
```
gemcitabine_line=data.frame("Gemcitabine drug sensitivity"= drugsensitivity[1:61,8])
everything=cbind(Cell_Line,gemcitabine_line)
plot(everything) #Quite interesting to look at
```{r}
gemcitabine_line=data.frame("Gemcitabine drug sensitivity"= drugsensitivity[1:61,8])
everything=cbind(Cell_Line,gemcitabine_line)

library(ggplot2)
basic=ggplot(everything, aes(Cell.Line.name, Gemcitabine.drug.sensitivity, colour = factor(Microsatellite), shape = factor(Microsatellite) )) +
   geom_point()
basic + ggtitle("Drug Sensitivity of Gemcitabine treated cell lines according to High Microsatellite Instability")+ xlab("Cell Line")+ylab("Drug Sensitivity")+theme(axis.text.x = element_text(angle = 90))
basic #very satisfied with this graph

# a relationship between high drug sensitivity(Gemcitabine) and high microsatellite instability cannot be concluded.
```

#Heatmap of biomarkers and of MSI related genes 
```{r}
#check which rows are for gemcitibine
gmci <- grepl('gemcitibine', colnames(fcgeneexpression))

#rows 365:420, make dataframe with gemci and biomarkers
biomarkers_geneexpression <- fcgeneexpression[c("ATF3", "CXCL8", "GADD45A", "CDKN1A", "TNFAIP3", "EGR1", "GDF15", "BTG2", "GEM", "CEACAM1", "DHRS2", "HBEGF", "IL11", "TNFRSF9", "TXNIP"),365:420]

#MSI related genes, choosen after research, new matrix for untreated and FC
geneexpression_MSI_related <- fcgeneexpression[c("MLH3", "MLH1", "MSH6", "PMS2CL///PMS2", "BATF", "BATF3"), 365:420]
geneexpression_MSI_related_untreated <- untreated_normalized[c("MLH3", "MLH1", "MSH6", "PMS2CL///PMS2", "BATF", "BATF3"), 365:420]

#make colnames equal to celllines
colnames(biomarkers_geneexpression) = metadata[365:420, 2]
colnames(geneexpression_MSI_related) <- metadata[365:420,2]
colnames(geneexpression_MSI_related_untreated) <- metadata[365:420,2]

#which cellines MSI-H or L? Note them 
#View(Cell_Line)

#MSI-H in col = 8,12,11,17,19,27,42,45 of geneexpression MSI related which is only gemcitibine specific


#define new matrices to divide matrices in MSI-H and MSI-L

biomarkers_geneexpression_MSIH <- biomarkers_geneexpression[,c(8,12,11,17,19,27,42,45)]
biomarkers_geneexpression_MSIL <- biomarkers_geneexpression[,-c(8,12,11,17,19,27,42,45)]

geneexpression_MSI_related_MSIH <- geneexpression_MSI_related[,c(8,12,11,17,19,27,42,45)]
geneexpression_MSI_related_MSIL <- geneexpression_MSI_related[,-c(8,12,11,17,19,27,42,45)]

geneexpression_MSI_related_untreated_MSIH <- geneexpression_MSI_related_untreated[, c(8,12,11,17,19,27,42,45)]
geneexpression_MSI_related_untreated_MSIL <- geneexpression_MSI_related_untreated[, -c(8,12,11,17,19,27,42,45)]


#packages which makes it possible to plot tow heatmaps next to each other
```
```{r echo=TRUE, results = 'hide', warning=FALSE, message = F}
library("ComplexHeatmap")
library("circlize")
library("colorspace")
library("GetoptLong")
```
Does MSI status effect the expression of biomarkers after gemcitibine treatment?
```{r}
#two heatmaps of FC of biomarkers in MSI-Low and MSI-High celllines
ht_biomarkers_geneexpression_MSIH = Heatmap(biomarkers_geneexpression_MSIH, name = "FC MSI-High", column_title = "MSI-H", column_names_gp = gpar(fontsize = 7)) 
ht_biomarkers_geneexpression_MSIL = Heatmap(biomarkers_geneexpression_MSIL, name = "FC MSI-Low", column_title = "MSI-L", column_names_gp = gpar(fontsize = 7))
 
ht_biomarkers_geneexpression_MSIH + ht_biomarkers_geneexpression_MSIL


#no obvious difference between MSI-H and MSI-L when it comes to FC after gemci treatment
#sum mean of FC in MSI-L (48 celllines) and MSI-H (8 celllines)

sum(biomarkers_geneexpression_MSIH)/8 ; sum(biomarkers_geneexpression_MSIL)/48
t.test(biomarkers_geneexpression_MSIH, biomarkers_geneexpression_MSIL, paired = TRUE)

#no significant difference between geneexpression in MSI-H and MSI-L biomarkers

#######Plotting mutations######
#compare mean of mutations, biomarkers_geneexpression_MSIH and L has to be changed, MSIH contains only MSIH celllines
mean_mutation_MSIH <- sum(mutations[,16] %in% colnames(biomarkers_geneexpression_MSIH))/8
mean_mutation_MSIL <- sum(mutations[,16] %in% colnames(biomarkers_geneexpression_MSIL))/48
```
```{r}
#is the drug sensitivity for gemcitabine dependend on the cell line? 
#plotting mean graph plot 

#calculation of GI50 mean for gemcitabine
rowMeans(drugsensitivity, na.rm = TRUE, dims = 1)
#GI50 mean value is 7,003322 for gemcitabine
#creating a vector for GI50 values of gemcitabine only 
NegLogGI50gemcitabine<- drugsensitivity[-1:-7,]
NegLogGI50gemcitabine2<- NegLogGI50gemcitabine[-2:-8,]

#substract the value 7,003322 for GI50 value of gemcitabine for every cell line 
NegLogGI50gemcitabine3 <- NegLogGI50gemcitabine2 - 7.003322
Neg_LogGI50gemcitabine4 <- as.data.frame(NegLogGI50gemcitabine3)


library(ggplot2)
theme_set(theme_bw())  
Neg_LogGI50gemcitabine4$`cell_line` <- rownames(Neg_LogGI50gemcitabine4)
Neg_LogGI50gemcitabine4$NegLogGI50gemcitabine3_type<- ifelse(Neg_LogGI50gemcitabine4$NegLogGI50gemcitabine3< 0, "below", "above")
ggplot(Neg_LogGI50gemcitabine4, aes(x=`cell_line`, y=NegLogGI50gemcitabine3, label=NegLogGI50gemcitabine3)) + 
    geom_bar(stat='identity', aes(fill=NegLogGI50gemcitabine3_type), width=2.5)  +
    scale_fill_manual(name="value", 
                       labels = c("Above Average", "Below Average"), 
                       values = c("above"="#00ba38", "below"="#f8766d")) + 
     labs(subtitle=" ", 
          title= "cell line and drug sensitivity") + 
coord_flip()

#looking at the previous mean graph plot now compare the differences in mutations of 10 cell lines which had high gemcitabine sensitivity and which had low sensitivity   

sum(mutations[,16] == "HS-578T")
#the cellline appears in the first 267 rows

#check the number of a certain mutation Typ (missense, nonsense, silent, slpice site) for the cell line HS-578T 
#create a matrix for each cell line

HS578T <- cbind(sum(mutations[1:267,8] == "Missense_Mutation"), sum(mutations[1:267,8] == "Silent"), sum(mutations[1:267,8] == "Nonsense_Mutation"),sum(mutations[1:267,8] == "Splice_Site"))

#unsure if to proceed for all 60 cell lines or only cell lines with top ten highest/lowest GI50

``` 

## Linear regression Analysis

(Still in progress, but would happily accept some feedback)

The goal of the linear regression is to find a way to predict the drug sensitivity (GI50 value) of celllines, based on their normal gene expression (when untreated) and/or based on their gene copy number. To do this, we will start by making sure that both matrixes (untreated and copy number) have the same rows and columns (same genes and same celllines) to be able to make a reasonable comparison.

Check dimension of untreated_normalized and copynumber to see if they are the same

```{r}
dim(untreated_normalized)
dim(copynumber)
```

Dimensions are completely different:
1. copynumber lists more genes than untreated_normalized
Solution: 
  - Check if rows (genes) are ordered alphabetically
  - If not: order both matrixes alphabetically by row name
  - Delete rows that appear in only one of the matrixes

2. untreated_normalized has more columns, since there is one column for each drug
 Solution:
 - New matrix with the mean for each cellline in untreated_normalized
 - Order columns alphabetically in both matrixes if not already ordered

#### Solving problem 1:

COPY NUMBER:

```{r}
identical(rownames(copynumber), rownames(copynumber[order(rownames(copynumber)),]))
```
False, so order them:
```{r}
cnforlr <- copynumber[order(rownames(copynumber)),]
```

UNTREATED: 

```{r}
identical(rownames(untreated_normalized), rownames(untreated_normalized[order(rownames(untreated_normalized)),]))
```
False, so order them:
```{r}
utnforlr <- untreated_normalized[order(rownames(untreated_normalized)),]
```

Keeping only the rows (genes) that are in both matrixes (copy number and untreated):
```{r}
commongenes <- intersect(rownames(cnforlr), rownames(utnforlr))
cnforlr <- cnforlr[which(rownames(cnforlr) %in% commongenes),]
utnforlr <- utnforlr[which(rownames(utnforlr) %in% commongenes),]
```
#### Solving problem 2:

(Still pending)
